A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-26 tidyindex.qmd
an overview paragraph for the examples and example titles

To construct an index, experts typically first define an underlying concept of interest to measure. This concept is latent and often without direct measure but of social and public importance. Relevant and available variables are then defined and collected for measuring the concept and combined as an index using statistical methods. Many decisions are factored into the construction process of an index. For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes. Indexes using weighted linear combination of variables, for example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita), needs to decide on the weights of each variable. In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data. All these decisions can introduce uncertainty to the index and lead to different interpretation and actions in operation.

@@

## A drought index example

@@

## Global Gender Gap Index

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights. The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index. The four dimensions are then aggregated in another linear combination with equal weight tp obtain the index. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.

@@

```{r eval = FALSE}
dt <- tidyindex::gggi %>% 
  dplyr::select(country: political_empowerment) %>%
  filter(region %in% c("South Asia", "Sub-Saharan Africa")) %>%
  arrange(index) 
colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")

>>>>

To construct an index, experts typically first define an underlying concept of interest to measure. This concept is latent and often without direct measure but of social and public importance. Relevant and available variables are then defined and collected for measuring the concept and combined as an index using statistical methods. Many decisions are factored into the construction process of an index. For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes. In indexes constructed from linear combination of variables, decisions need to be made on the weight assigned to each variable. In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data. All these decisions can introduce uncertainty to the index and lead to different interpretation and actions in operation.

@@

This section uses the example of drought and social indexes to show the analysis made possible with the index pipeline. The drought index example computes two indexes with various time scales and distributions simultaneously using the pipeline framework to understand the flood and drought events in Queensland. The social index example focuses on the dimension reduction in Global Gender Gap Index to explore the impact of weight changes in linear combination on index value and country ranking.

## Every distribution, every scale, every index all at once

@@

## Does a minor change in variable weights cause a tornado?

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights. The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index. The four dimensions are then aggregated in another linear combination with equal weight to obtain the index. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.

@@

```{r eval = FALSE}
dt <- tidyindex::gggi %>% 
  dplyr::select(country: political_empowerment) %>%
  filter(region %in% c("Sub-Saharan Africa")) %>%
  arrange(index) 
colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")

### 2023-07-26 tidyindex.tex
an overview paragraph for the examples and example titles

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

To construct an index, experts typically first define an underlying
concept of interest to measure. This concept is latent and often without
direct measure but of social and public importance. Relevant and
available variables are then defined and collected for measuring the
concept and combined as an index using statistical methods. Many
decisions are factored into the construction process of an index. For
example, decision on which variables to include may depend on the data
availability and the choice in other similar indexes. Indexes using
weighted linear combination of variables, for example, Human Development
Index (HDI) (United Nations Development Programme 2022) combines the
three aspects (life expectancy, years of schooling, and GNI per capita),
needs to decide on the weights of each variable. In the case of a
drought index, analysts may need to decide among similar indexes, on the
time scale to aggregate precipitation, and the distribution used to fit
the data. All these decisions can introduce uncertainty to the index and
lead to different interpretation and actions in operation.

@@

\hypertarget{a-drought-index-example}{%
\subsection{A drought index example}\label{a-drought-index-example}}

@@

\hypertarget{global-gender-gap-index}{%
\subsection{Global Gender Gap Index}\label{global-gender-gap-index}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
United Nations Development Programme. 2022. {``Human Development Report
2021-22.''} New York. \url{http://report.hdr.undp.org}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

To construct an index, experts typically first define an underlying
concept of interest to measure. This concept is latent and often without
direct measure but of social and public importance. Relevant and
available variables are then defined and collected for measuring the
concept and combined as an index using statistical methods. Many
decisions are factored into the construction process of an index. For
example, decision on which variables to include may depend on the data
availability and the choice in other similar indexes. In indexes
constructed from linear combination of variables, decisions need to be
made on the weight assigned to each variable. In the case of a drought
index, analysts may need to decide among similar indexes, on the time
scale to aggregate precipitation, and the distribution used to fit the
data. All these decisions can introduce uncertainty to the index and
lead to different interpretation and actions in operation.

@@

This section uses the example of drought and social indexes to show the
analysis made possible with the index pipeline. The drought index
example computes two indexes with various time scales and distributions
simultaneously using the pipeline framework to understand the flood and
drought events in Queensland. The social index example focuses on the
dimension reduction in Global Gender Gap Index to explore the impact of
weight changes in linear combination on index value and country ranking.

\hypertarget{every-distribution-every-scale-every-index-all-at-once}{%
\subsection{Every distribution, every scale, every index all at
once}\label{every-distribution-every-scale-every-index-all-at-once}}

@@

\hypertarget{does-a-minor-change-in-variable-weights-cause-a-tornado}{%
\subsection{Does a minor change in variable weights cause a
tornado?}\label{does-a-minor-change-in-variable-weights-cause-a-tornado}}

@@
